---
title: "Make a ternary plot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Make a ternary plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To create a ternary plot from origin and insertion data, first we need data. Attachment centroid data are available for the AL_008 specimen (although similar data can be extracted directly from stl files; see the "Working with stl files" vignette).

```{r, warning=FALSE, message=FALSE}
library(MuscleTernary)

AL008 <- read_csv(system.file("extdata",
                              "AL_008_data.csv",
                              package = "MuscleTernary"),
                  show_col_types = FALSE) |>
  dplyr::select(-side)
AL008
```

We need to convert the x, y, and z origin and insertion columns into relative ternary coordinates using the `coords_to_ternary()` function. Here we also use the `grouping` option to return the mean for each muscle.

```{r}
ternary_coords <- coords_to_ternary(coords = AL008, grouping = c("muscle"))
ternary_coords
```

The `force` column contains the estimated force for each muscle. Force can be estimated from PCSA, which can be derived from attachment areas (see the "Working with stl files" vignette).

The simplest ternary plot simply maps the `x`, `y`, and `z` columns and colors points by muscle:

```{r}
ggtern(data = ternary_coords,
       aes(x = x, y = y, z = z,
           color = muscle)) +
  geom_point()
```

We can make this a more ready-to-use plot by recoloring using the standard palette for muscle colors. `muscle_color_map()` is a function that returns the mapping of muscle to its hex code.

```{r}
muscle_color_map

ggtern(data = ternary_coords,
       aes(x = x, y = y, z = z,
           color = muscle,
           size = force)) +
  geom_point() +
  muscle_color_map()
```

And finally we can add labels, change the theme, and update the scale, resizing the points according to force.

```{r}
ggtern(data = ternary_coords,
       aes(x = x, y = y, z = z,
           color = muscle,
           size = force)) +
  geom_point() +
  muscle_color_map() +
  labs(x       = "ML",
       xarrow  = "Mediolateral",
       y       = "DV",
       yarrow  = "Dorsoventral",
       z       = "RC",
       zarrow  = "Rostrocaudal") +
  theme_bw(base_size = 16) +
  theme_showarrows() +
  scale_size_continuous(range = c(5, 15), name = "Force (N)") +
  guides(colour = guide_legend(override.aes = list(size = 6),
                               ncol = 2, byrow = TRUE))
```
